Data analysis to prioritise new cycleways in response to COVID-19

### 2020-06-03 report.Rmd
Close #71, fix #72
Update report

The £250m released by the Emergency Active Travel Fund will be spent in three broad ways:

@@

The approach was designed to identify roads and road segments where investment in new cycleways should be prioritised using a transparent methodology and nationally consistent datasets, alongside local authority expertise and local knowledge.
Strong and actionable datasets already exist to support strategic cycle network planning, notably the Propensity to Cycle Tool (PCT), a national planning support system [@lovelace_propensity_2017].
The PCT is already used by local authorities and contains key data showing not only current levels of cycling but also scenarios of growth, notably the Government Target scenario of doubling the number of cycling stages made by 2025 [@departmentfortransport_cycling_2017], at area, route and route network levels, for commuting and school travel (as of May 2020) [@goodman_scenarios_2019].
A limitation of the PCT, however, is that it takes little account of existing infrastructure: variables such as road width and even the presence/absence of cycle infrastructure currently are not included in the results.

To address this limitation, the work set out below will use national road network data which includes geographic attributes and cycle infrastructure from OpenStreetMap, plus carriageway width data derived from analysis of Ordnance Survey data.
This road data was developed for the Cycling Infrastructure Prioritisation Tool (CyIPT) project, funded as a prototype project by a DfT Transport-technology research innovation grants (T-TRIG).
The results of CyIPT, while not publicly accessible, are well known among local authority transport planners and are reported to be of great use to practitioners in scheme prioritisation, according to a DfT funded Discovery project on tools to support investment in active transport.^[
Planning Tools for cycle networks and infrastructure prioritisation, unpublished.
]

The rapid cycleway prioritisation tool is based on levels of cycling potential *and* road attribute data, with an emphasis on road space reallocation.
The method was parameterised to highlight roads that have sufficient space to allow for new cycleways without affecting the navigable network for motorised modes [@lovelace_methods_2020].
A key benefit of this method is that it allows continuous road sections to be identified and ranked, supporting prioritisation of investment, based not only on cycling potential, but also based on continuous road length and estimated cycle usage measured in km per day.
The tool highlights continuous road sections with high cycling potential in the context of existing infrastructure on the basis that new infrastructure should maximise cycle network cohesion and continuity [@parkin_designing_2018].

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, out.width = "100%", eval = FALSE)
```

@@

To identify road sections on which there is a spare lane we developed a simple algorithm that takes the OSM variable [`lanes`](https://wiki.openstreetmap.org/wiki/Key:lanes) if it is present and, if not, derives the number from the highway type and presence/absence of bus lanes.
All segments defined as having a spare lane using this method is shown in Figure \@ref(fig:levels) (left).
In future, this methodology could be enhanced to take into account the effect of lane widths, which increase the effective available space, and the presence of vehicle parking bays, which reduce it.
The result of filtering by distance and cycling potential before and after grouping using graph membership of touching roads is shown in \@ref(fig:levels) (middle and right, respectively).
Grouping linked roads before filtering results in a more cohesive network.

<!-- Note this could be a function in an R package.. -->
<!-- see https://github.com/cyipt/cyipt/blob/82248b2f99e388fac314d34ec5aa49bb90a737a3/scripts/prep_data/clean_osm.R#L349 -->

<!-- reallocated road space in particular -->
<!-- An important distinction when developing methods for automated analysis of transport networks is the level of analysis. -->

```{r levels, fig.height=3, fig.cap="Illustration of the 'group then filter' method to identify long sections with spare lanes *and* high cycling potential"}
r_pct_lanes_all = r_central %>% 
  filter(cycling_potential > min_cycling_potential) %>% 
  filter(lanes_f > 1 | lanes_b > 1)
# mapview::mapview(r_pct_lanes)

touching_list = st_touches(r_pct_lanes_all)
g = igraph::graph.adjlist(touching_list)
components = igraph::components(g)
r_pct_lanes_all$group = components$membership
r_pct_lanes = r_pct_lanes_all %>% 
  group_by(group) %>% 
  mutate(group_length = sum(length)) %>% 
  mutate(cycling_potential_mean = weighted.mean(cycling_potential, w = length, na.rm = TRUE)) %>% 
  filter(cycling_potential_mean > min_grouped_cycling_potential)
r_pct_lanes = r_pct_lanes %>% filter(group_length > min_grouped_length)

r_pct_lanes$graph_group = r_pct_lanes$group
group_table = table(r_pct_lanes$group)
top_groups = tail(sort(group_table), 5)
r_pct_lanes$graph_group[!r_pct_lanes$graph_group %in% names(top_groups)] = "other"

r_filter_before_grouping = rj %>% 
  filter(cycling_potential > min_cycling_potential) %>% 
  filter(lanes_f > 1 | lanes_b > 1) %>% 
  filter(cycling_potential > min_grouped_cycling_potential) %>% 
  filter(length > 100)
tmap_mode("plot")
m0 = tm_shape(city_key_buffer) + tm_borders(col = "grey") +
  tm_shape(r_pct_lanes_all) + tm_lines() +
  tm_layout(title = "Roads on which there are spare lanes.")
m1 = tm_shape(city_key_buffer) + tm_borders(col = "grey") +
  tm_shape(r_filter_before_grouping) + tm_lines() +
  tm_layout(title = "Filter then group:\n(length > 100, cycling_potential > 100)")
m2 = tm_shape(city_key_buffer) + tm_borders(col = "grey") +
  tm_shape(r_pct_lanes) + tm_lines("graph_group", palette = "Dark2") +
  tm_layout(title = "Group then filter:\n(length > 500, cycling_potential > 100)")
# todo show group membership with colours
ma = tmap_arrange(m0, m1, m2, nrow = 1)
ma
```

```{r joining}
r_pct_grouped = r_pct_lanes %>%
  group_by(name, group) %>%
  summarise(
    group_length = sum(length),
    cycling_potential = round(weighted.mean(cycling_potential_mean, length))
  )
# summary(r_pct_grouped$group_length)
r_pct_top = r_pct_grouped %>%
  filter(group_length > min_grouped_length) %>% 
  filter(cycling_potential > min_grouped_cycling_potential) %>% 
  filter(!grepl(pattern = regexclude, name, ignore.case = TRUE)) %>% 
  mutate(km_day = round(cycling_potential * group_length / 1000))
# head(r_pct_top$km_day)

r_pct_lanes_overlap = r_pct_lanes[r_pct_top, , op = sf::st_covered_by] # works
# r_pct_no_overlap = st_difference(r_pct_lanes, r_pct_top) # slow
r_pct_no_overlap = r_pct_lanes %>% 
  filter(!idGlobal %in% r_pct_lanes_overlap$idGlobal)

r_pct_top_n = r_pct_top %>% 
  group_by(name) %>% 
  slice(which.max(km_day)) %>% 
  filter(name != "") %>% 
  ungroup() %>% 
  top_n(10, wt = km_day) 
# aim: make the labels go on the longest sections
# %>% 
#   st_cast("LINESTRING") %>% 
#   mutate(length_section = as.numeric(st_length(.))) %>% 
#   group_by(name) %>% 
#   slice(which.max(length_section))
# r_pct_top_n$name
# summary(r_pct_grouped$group_length)
```

>>>>

The £250m released by the Emergency Active Travel Fund will be spent in three broad [ways](https://www.transportxtra.com/publications/local-transport-today/news/65661/dft-releases-first-45m-of-250m-fund/):

@@

The approach was designed to identify roads and road segments where investment in new cycleways should be prioritised using a transparent methodology and nationally consistent datasets, alongside local authority expertise and local knowledge.
Strong and actionable datasets already exist to support strategic cycle network planning, notably the Propensity to Cycle Tool (PCT), a national planning support system [@lovelace_propensity_2017].
The PCT is already used by local authorities and contains key data showing not only current levels of cycling but also scenarios of growth, notably the Government Target scenario of doubling the number of cycling stages made by 2025 [@departmentfortransport_cycling_2017], at area, route and route network levels, for commuting and school travel (as of May 2020) [@goodman_scenarios_2019].

A limitation of the PCT is that it takes little account of existing infrastructure: variables such as road width and even the presence/absence of cycle infrastructure currently are not included in the results.
The rapid cycleway prioritisation tool overcomes this issu by using road network data which includes geographic attributes and cycle infrastructure from OpenStreetMap.
Carriageway width data were taken from the Cycling Infrastructure Prioritisation Tool (CyIPT) project, funded as a prototype project by a DfT Transport-technology research innovation grants (T-TRIG).

The rapid cycleway prioritisation tool is based on levels of cycling potential *and* road attribute data, with an emphasis on road space reallocation.
The method was parameterised to highlight roads that have sufficient space to allow for new cycleways without affecting the navigable network for motorised modes [@lovelace_methods_2020].
A benefit of this method is that it allows continuous road sections to be identified and ranked, supporting prioritisation of investment.
Roads can be ranked based not only on cycling potential, but also based on continuous road length and estimated cycle usage measured in km per day.
The tool highlights continuous road sections with high cycling potential in the context of existing infrastructure on the basis that new infrastructure should maximise cycle network cohesion and continuity [@parkin_designing_2018].

@@

To identify road sections on which there is a spare lane we developed a simple algorithm that takes the OSM variable [`lanes`](https://wiki.openstreetmap.org/wiki/Key:lanes) if it is present and, if not, derives the number from the highway type and presence/absence of bus lanes.
All segments defined as having a spare lane using this method are shown in Figure \@ref(fig:levels) (left).
We also included estimates of road width, only flagging roads with an average estimated width of more than 10m (the absolute minumum that should be considered for cycleways without additional changes).
In future, this methodology could be enhanced to include verge and pavement width estimates, which could increase the effective available space for new cycleways, and the presence of vehicle parking bays, and on street parking which reduce it.
The result of filtering by distance and cycling potential before and after grouping using graph membership of touching roads is shown in Figure \@ref(fig:levels) (middle and right, respectively), which shows that grouping linked roads before filtering results in a more cohesive network.

```{r levels, out.width="90%", fig.cap="Illustration of the 'group then filter' method to identify long sections with spare lanes *and* high cycling potential"}
knitr::include_graphics("figures/filter-then-group.png")
```

## Feedback from stakeholders

We elicited feedback from stakeholders via the Basecamp forum for local authority transport planners.
As a result of the feedback we made a few adjustments to the results:

- We refined the road identification method, to highlight more central road sections and fewer long sections far from high density areas, in response to feedback suggesting that some of the road sections went too far into the countryside.
- We simplified the legend of the basemap, so that the layers were presented in time order: Existing cycleway (now); Top ranked new cycleways (near term); Cohesive network (long term).
- We requested data from Highways England to flag the recommendations on the Strategic Road Network (SRN)
